Reasonably Programmable Literal Notation

### 2018-07-08 icfp18/supplement/icfp18-supplement.tex
final submitted version of paper and supplement!

%\arpasupport{fox}
\support{This work was
supported in part by AFRL and DARPA under agreement \#FA8750-16-2-0042;
 by the NSA under lablet contract \#H98230-14-C-0140; and by Ravi Chugh
 at the University of Chicago.}
% \authorsupport{The author holds a Froboz Gradual Fellowship.}

>>>>

%\arpasupport{fox}
\support{This work was
supported in part by AFRL and DARPA under agreement \#FA8750-16-2-0042;
 by the NSA under lablet contract \#H98230-14-C-0140.}
% \authorsupport{The author holds a Froboz Gradual Fellowship.}

### 2018-07-08 icfp18/syntax-icfp18.tex
final submitted version of paper and supplement!

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission
% \setcopyright{none}             %% For review submission
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%%% The following is specific to ICFP'18 and the paper
%%% 'Reasonably Programmable Literal Notation'
%%% by Cyrus Omar and Jonathan Aldrich.
%%%
\setcopyright{rightsretained}
\acmPrice{}
\acmDOI{10.1145/3234629}
\acmYear{2018}
\copyrightyear{2018}
\acmJournal{PACMPL}
\acmVolume{2}
\acmNumber{ICFP}
\acmArticle{40}
\acmMonth{9}

@@

\input{thesis/macros}

@@

\begin{figure}[h]
% \vspace{-9px}
\begin{lstlisting}
  fun heading first_name = SURL<xml><h1>Hello, {[EURLfirst_nameSURL]}!</h1></xml>EURL
  val body = SURL<xml><body>{EURLheading "World"SURL} ...</body></xml>EURL
\end{lstlisting}
\vspace{-3px}
\caption{HTML literals with support for splicing at two types are built primitively into Ur/Web \cite{conf/popl/Chlipala15}.}
\label{fig:urweb}
\vspace{-10px}
\end{figure}

@@

This section presents  $\miniVersePat$, a calculus of simple expression and pattern TLMs based on ``core ML''. By the end of this section, we will have theorems that formalize the reasoning principles that were outlined informally in the previous sections. $\miniVersePat$ consists of an {unexpanded language} ({UL}) defined by typed expansion to the core language, named for the purposes of this paper $\miniVerseBase$.

@@

\subsection{Core Language}\label{sec:s-XL}
Fig. \ref{fig:U-expanded-terms} gives the syntax of the core language, $\miniVerseBase$. This language forms a standard pure functional language with partial function types, quantification over types, recursive types, labeled product types, labeled sum types and support for pattern matching.\footnote{The accompanying technical report shows how to remove pattern matching  (and pattern TLMs) from the calculus by adding the usual elimination forms, e.g. $\unfold{e}$ rather than the $\aefoldp{p}$ pattern for values of recursive type.} Formally, core language terms are {abstract binding trees} (ABTs) identified up to alpha-equivalence, so we follow the syntactic conventions of \citet{pfple1}. This notational convention also clearly distinguishes core terms from unexpanded terms, which formally behave quite differently as we will describe below.

 The reader is directed to \emph{PFPL} \cite{pfple1} for a detailed introductory account of all of these language constructs. We will only tersely summarize the static and dynamic semantics of the core language because the particularities are not critical to our ideas. The static semantics is organized around the type formation judgement, $\istypeU{\Delta}{\tau}$, the expression typing judgement, $\hastypeU{\Delta}{\Gamma}{e}{\tau}$, and the pattern typing judgement, $\patType{\pctx}{p}{\tau}$. Type formation contexts, $\Delta$, track hypothesis of the form $\Dhyp{t}$, and typing contexts, $\Gamma$, track hypotheses of the form $x : \tau$. In the pattern typing judgement, $\pctx$ collects the typing hypotheses generated by $p$. These judgements are inductively defined in the accompanying technical report along with necessary auxiliary structures and standard lemmas. The dynamic semantics of $\miniVerseBase$ is organized around the judgements $\isvalU{e}$, which says that $e$ is a value, and $\evalU{e}{e'}$, which says that $e$ evaluates to the value $e'$. As in ML, evaluation can diverge (general recursion is possible via recursive types \cite{pfple1}). It can also result in match failure.%\emph{Type formation contexts}, $\Delta$, and \emph{typing contexts}, $\Gamma$ are defined in the accompanying technical report in the standard way \cite{pfpl}, 
% \[\begin{array}{ll}
% % \textbf{Judgement Form} & \textbf{Description}\\
% \istypeU{\Delta}{\tau} & \text{$\tau$ is a well-formed type}\\
% %\isctxU{\Delta}{\Gamma} & \text{$\Gamma$ is a well-formed typing context assuming $\Delta$}\\
% \hastypeU{\Delta}{\Gamma}{e}{\tau} & \text{$e$ is assigned type $\tau$}\\
% \ruleType{\Delta}{\Gamma}{r}{\tau}{\tau'} & \text{$r$ takes values of type $\tau$ to $\tau'$}\\
% \patType{\pctx}{p}{\tau} & \text{$p$ matches values of type $\tau$}\\
% & \text{and generates hypotheses $\pctx$} 
% \end{array}\]
% \emph{Type formation contexts}, $\Delta$, are finite sets of hypotheses of the form $\Dhyp{t}$. %Empty finite sets are written $\emptyset$, or omitted entirely within judgements, and non-empty finite sets are written as comma-separated finite sequences identified up to exchange and contraction. We write $\Delta, \Dhyp{t}$ when $\Dhyp{t} \notin \Delta$ for $\Delta$ extended with the hypothesis $\Dhyp{t}$. %Finite sets are written as finite sequences identified up to exchange.% We write $\Dcons{\Delta}{\Delta'}$ for the union of $\Delta$ and $\Delta'$.
% \emph{Typing contexts}, $\Gamma$, are finite functions that map each variable $x \in \domof{\Gamma}$, where $\domof{\Gamma}$ is a finite set of variables, to the hypothesis $\Ghyp{x}{\tau}$, for some $\tau$. The judgements above are inductively defined in the accompanying technical reportal material and validate standard lemmas, also given in the accompanying technical report.
%Empty typing contexts are written $\emptyset$, or omitted entirely within judgements, and non-empty typing contexts are written as finite sequences of hypotheses identified up to exchange and contraction. We write $\Gamma, \Ghyp{x}{\tau}$, when $x \notin \domof{\Gamma}$, for the extension of $\Gamma$ with a mapping from $x$ to $\Ghyp{x}{\tau}$, and $\Gcons{\Gamma}{\Gamma'}$ when $\domof{\Gamma} \cap \domof{\Gamma'} = \emptyset$ for the typing context mapping each $x \in \domof{\Gamma} \cup \domof{\Gamma'}$ to $x : \tau$ if $x : \tau \in \Gamma$ or $x : \tau \in \Gamma'$. % We write $\isctxU{\Delta}{\Gamma}$ if every type in $\Gamma$ is well-formed relative to $\Delta$.
% \begin{definition}[Typing Context Formation] \label{def:isctxU}
% $\isctxU{\Delta}{\Gamma}$ iff for each hypothesis $x : \tau \in \Gamma$, we have $\istypeU{\Delta}{\tau}$.
% \end{definition}

@@

\begin{figure}
%\hspace{-5px}
\begin{minipage}{\textwidth}
\small
$\arraycolsep=2pt\begin{array}{llcl}
\mathsf{Typ} & \tau & ::= & t ~\vert~ \aparr{\tau}{\tau} ~\vert~ \aall{t}{\tau} ~\vert~ \arec{t}{\tau} ~\vert~ \aprod{\labelset}{\mapschema{\tau}{i}{\labelset}} ~\vert~ \asum{\labelset}{\mapschema{\tau}{i}{\labelset}}\\
\mathsf{Exp} & e & ::= & x ~\vert~ \aelam{\tau}{x}{e} ~\vert~ \aeap{e}{e} ~\vert~ \aetlam{t}{e} ~\vert~ \aetap{e}{\tau} ~\vert~ \aefold{e} ~\vert~ \aetpl{\labelset}{\mapschema{e}{i}{\labelset}} ~\vert~  \aein{\ell}{e} \\
& & \vert & \aematchwith{n}{e}{\seqschemaX{r}}\\
\mathsf{Rule} & r & ::= & \aematchrule{p}{e}\\
\mathsf{Pat} & p & ::= & x  ~\vert~ \aewildp ~\vert~ \aefoldp{p} ~\vert~ \aetplp{\labelset}{\mapschema{p}{i}{\labelset}} ~\vert~ \aeinjp{\ell}{p}
\end{array}$
\end{minipage}
\caption[Syntax of the core language of $\miniVersePat$]{Syntax of the core language, $\miniVerseBase$, which is an entirely standard typed lambda calculus.  Metavariable $x$ ranges over variables, $t$ over type variables, $\ell$ over labels and $L$ over finite sets of labels. We write $\mapschema{\tau}{i}{\labelset}$ for a finite mapping of each label $i$ in $\labelset$ to some type $\tau_i$, and similarly for other sorts of terms. We write $\seqschemaX{r}$ for a finite sequence of $n > 0$  rules. %When using stylized forms, the label set is omitted when it can be inferred, e.g. the labeled product type $\prodt{\finmap{\mapitem{\ell_1}{e_1}, \mapitem{\ell_2}{e_2}}}$ leaves the label set $\{\ell_1, \ell_2\}$ implicit. 
% When we use the stylized forms, we assume that the reader can infer suppressed indices and arguments from the surrounding context.
}
\label{fig:U-expanded-terms}
\end{figure}

@@

\paragraph{Common Forms} Most of the proto-expansion forms, including all of those elided in Fig.~\ref{fig:U-candidate-terms} mirror corresponding expanded forms. The rules governing proto-expansion validation for these common forms  in the accompanying technical report correspondingly mirror the typing rules. Splicing scenes---$\escenev$, $\tscenev$ and $\pscenev$--pass opaquely through these rules, i.e. none of these rules can access the application site contexts. Notice that the initial expansion-interal typing contexts, $\Delta$ and $\Gamma$, start out empty in the rules in Fig.~\ref{fig:ee-ap-setlm}. This maintains context independence (defined formally below).

@@

\paragraph{TLM Definitions} The Reason grammar extension turns TLM definitions into module definitions of a stereotyped form. For example, definition of \li{$regex} from Fig.~\ref{fig:regex-tlm-def} is, eliding a few minor details, implemented as shown below. The comments describe how the components correspond.

\begin{lstlisting}
    module RelitInternalDefn_regex = struct
      type t = Regex.t (* corresponds to "at Regex.t" *)
      module Lexer_RegexLexer = struct end (* ... "lexer RegexLexer" *)
      module Parser_RegexParser = struct end (* ... "parser RegexParser" *)
      module Nonterminal_start = struct end (* ... ".start" *)
      module Package_regex_parser = struct end (* ... "in package regex_parser" *)
      module Dependencies = struct (* ... "dependencies = {" *)
        module Regex = Regex
      end
      exception Apply of string * string (* used internally, see below *)
    end
\end{lstlisting}

The name of the TLM is stored as the module name, with the internal prefix \li{RelitInternalDefn_}. By encoding TLM definitions as modules, TLMs can be packaged inside other modules (\li{RegexNotation} in Fig.~\ref{fig:regex-tlm-def}) and it is straightforward to support scoped TLM abbreviations, as on Line 1 of Fig.~\ref{fig:first-tlm-example}. By encoding the relevant information from the TLM definition in module names, e.g. \li{Lexer_RegexLexer} and \li{Nonterminal_start}, we ensure that the corresponding signature is a \emph{singleton signature}, i.e. it uniquely identifies the TLM definition, so TLM definitions can be included in module signatures, exported from functors and packaged using the standard tooling.

@@

The Relit preprocessor rewrites TLM applications by following the specification in the previous section, differing only in that the ``dependency variable'' is a ``dependency path'', here to \li{Notation_regex___relit.Dependencies}. The only difficult is in finding the TLM definition at all, which due to our integration into the module system may have been accessed indirectly, e.g. here via the \li{open} directive and in other situations via module synonyms, functors and so on. The trick is in the singleton restriction just described---if we can determine the signature for the module path that \li{Apply} is accessed from, we have the full definition in hand. To do so, we run the OCaml typechecker on this unexpanded representation. Because the TLM application raises an exception, its type can be generalized arbitrarily. As long as unrelated type errors at a particular depth are resolved (treating the TLM applications at that depth as ``holes'') expansion at that depth can proceed. This occasionally causes types that incidentally appear in error messages to be more general than the programmer might expect (e.g. here \li{restriction_template} will have type \li{'a => 'b} during this ``pre-typing'' phase). 
%If this type ends up in an error message incidentally, e.g. due to a signature mismatch elsewhere, it may be somewhat confusing. 
We are working to improve error reporting and, more generally, improving integration with various other tools in the Reason / OCaml ecosystem.

@@

Despite these semantic constraints, the system is able to express a number of non-trivial examples with few compromises because there is only one simple lexical constraint on literal bodies. The mechanism integrates cleanly into Reason/OCaml, with full support for its type system, module system and pattern matching system. TLM providers can use existing, mature parser generators and parse tree representations, and can opt-in to a stronger typing discipline by using various implementations of MetaOCaml. Overall, we believe that TLMs represent a distinctly \emph{reasonable} and \emph{expressive} new point in the design space of literal notation definition systems.
\vspace{-5px}

@@

\clearpage
%% Bibliography
\bibliography{../../../papers/research}

>>>>

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission
% \setcopyright{none}             %% For review submission
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%%% The following is specific to ICFP'18 and the paper
%%% 'Reasonably Programmable Literal Notation'
%%% by Cyrus Omar and Jonathan Aldrich.
%%%
\setcopyright{rightsretained}
\acmPrice{}
\acmDOI{10.1145/3236801}
\acmYear{2018}
\copyrightyear{2018}
\acmJournal{PACMPL}
\acmVolume{2}
\acmNumber{ICFP}
\acmArticle{106}
\acmMonth{9}

@@

\input{macros}

@@

\begin{figure}[h]
\vspace{-4px}
\begin{lstlisting}
  fun heading first_name = SURL<xml><h1>Hello, {[EURLfirst_nameSURL]}!</h1></xml>EURL
  val body = SURL<xml><body>{EURLheading "World"SURL} ...</body></xml>EURL
\end{lstlisting}
\vspace{-3px}
\caption{HTML literals with support for splicing at two types are built primitively into Ur/Web \cite{conf/popl/Chlipala15}.}
\label{fig:urweb}
\vspace{-5px}
\end{figure}

@@

This section presents  $\miniVersePat$, a calculus of simple expression and pattern TLMs based on ``core ML''. By the end of this section, we will have theorems that formalize the reasoning principles that were outlined informally in the previous sections. $\miniVersePat$ consists of an {unexpanded language} ({UL}) defined by typed expansion to the core language, named for the purposes of this paper $\miniVerseCore$.

@@

\subsection{Core Language}\label{sec:s-XL}
Fig. \ref{fig:U-expanded-terms} gives the syntax of the core language, $\miniVerseCore$. This language forms a standard pure functional language with partial function types, quantification over types, recursive types, labeled product types, labeled sum types and support for pattern matching.\footnote{The accompanying technical report shows how to remove pattern matching  (and pattern TLMs) from the calculus by adding the usual elimination forms, e.g. $\unfold{e}$ rather than the $\aefoldp{p}$ pattern for values of recursive type.} Formally, core language terms are {abstract binding trees} (ABTs) identified up to alpha-equivalence, so we follow the syntactic conventions of \citet{pfple1}. This notational convention also clearly distinguishes core terms from unexpanded terms, which formally behave quite differently as we will describe below.

 The reader is directed to \emph{PFPL} \cite{pfple1} for a detailed introductory account of all of these language constructs. We will only tersely summarize the static and dynamic semantics of the core language because the particularities are not critical to our ideas. The static semantics is organized around the type formation judgement, $\istypeU{\Delta}{\tau}$, the expression typing judgement, $\hastypeU{\Delta}{\Gamma}{e}{\tau}$, and the pattern typing judgement, $\patType{\pctx}{p}{\tau}$. Type formation contexts, $\Delta$, track hypothesis of the form $\Dhyp{t}$, and typing contexts, $\Gamma$, track hypotheses of the form $x : \tau$. In the pattern typing judgement, $\pctx$ collects the typing hypotheses generated by $p$. These judgements are inductively defined in the accompanying technical report along with necessary auxiliary structures and standard lemmas. The dynamic semantics of $\miniVerseCore$ is organized around the judgements $\isvalU{e}$, which says that $e$ is a value, and $\evalU{e}{e'}$, which says that $e$ evaluates to the value $e'$. As in ML, evaluation can diverge (general recursion is possible via recursive types \cite{pfple1}). It can also result in match failure.%\emph{Type formation contexts}, $\Delta$, and \emph{typing contexts}, $\Gamma$ are defined in the accompanying technical report in the standard way \cite{pfpl}, 
% \[\begin{array}{ll}
% % \textbf{Judgement Form} & \textbf{Description}\\
% \istypeU{\Delta}{\tau} & \text{$\tau$ is a well-formed type}\\
% %\isctxU{\Delta}{\Gamma} & \text{$\Gamma$ is a well-formed typing context assuming $\Delta$}\\
% \hastypeU{\Delta}{\Gamma}{e}{\tau} & \text{$e$ is assigned type $\tau$}\\
% \ruleType{\Delta}{\Gamma}{r}{\tau}{\tau'} & \text{$r$ takes values of type $\tau$ to $\tau'$}\\
% \patType{\pctx}{p}{\tau} & \text{$p$ matches values of type $\tau$}\\
% & \text{and generates hypotheses $\pctx$} 
% \end{array}\]
% \emph{Type formation contexts}, $\Delta$, are finite sets of hypotheses of the form $\Dhyp{t}$. %Empty finite sets are written $\emptyset$, or omitted entirely within judgements, and non-empty finite sets are written as comma-separated finite sequences identified up to exchange and contraction. We write $\Delta, \Dhyp{t}$ when $\Dhyp{t} \notin \Delta$ for $\Delta$ extended with the hypothesis $\Dhyp{t}$. %Finite sets are written as finite sequences identified up to exchange.% We write $\Dcons{\Delta}{\Delta'}$ for the union of $\Delta$ and $\Delta'$.
% \emph{Typing contexts}, $\Gamma$, are finite functions that map each variable $x \in \domof{\Gamma}$, where $\domof{\Gamma}$ is a finite set of variables, to the hypothesis $\Ghyp{x}{\tau}$, for some $\tau$. The judgements above are inductively defined in the accompanying technical reportal material and validate standard lemmas, also given in the accompanying technical report.
%Empty typing contexts are written $\emptyset$, or omitted entirely within judgements, and non-empty typing contexts are written as finite sequences of hypotheses identified up to exchange and contraction. We write $\Gamma, \Ghyp{x}{\tau}$, when $x \notin \domof{\Gamma}$, for the extension of $\Gamma$ with a mapping from $x$ to $\Ghyp{x}{\tau}$, and $\Gcons{\Gamma}{\Gamma'}$ when $\domof{\Gamma} \cap \domof{\Gamma'} = \emptyset$ for the typing context mapping each $x \in \domof{\Gamma} \cup \domof{\Gamma'}$ to $x : \tau$ if $x : \tau \in \Gamma$ or $x : \tau \in \Gamma'$. % We write $\isctxU{\Delta}{\Gamma}$ if every type in $\Gamma$ is well-formed relative to $\Delta$.
% \begin{definition}[Typing Context Formation] \label{def:isctxU}
% $\isctxU{\Delta}{\Gamma}$ iff for each hypothesis $x : \tau \in \Gamma$, we have $\istypeU{\Delta}{\tau}$.
% \end{definition}

@@

\begin{figure}
%\hspace{-5px}
\begin{minipage}{\textwidth}
\small
$\arraycolsep=2pt\begin{array}{llcl}
\mathsf{Typ} & \tau & ::= & t ~\vert~ \aparr{\tau}{\tau} ~\vert~ \aall{t}{\tau} ~\vert~ \arec{t}{\tau} ~\vert~ \aprod{\labelset}{\mapschema{\tau}{i}{\labelset}} ~\vert~ \asum{\labelset}{\mapschema{\tau}{i}{\labelset}}\\
\mathsf{Exp} & e & ::= & x ~\vert~ \aelam{\tau}{x}{e} ~\vert~ \aeap{e}{e} ~\vert~ \aetlam{t}{e} ~\vert~ \aetap{e}{\tau} ~\vert~ \aefold{e} ~\vert~ \aetpl{\labelset}{\mapschema{e}{i}{\labelset}} ~\vert~  \aein{\ell}{e} \\
& & \vert & \aematchwith{n}{e}{\seqschemaX{r}}\\
\mathsf{Rule} & r & ::= & \aematchrule{p}{e}\\
\mathsf{Pat} & p & ::= & x  ~\vert~ \aewildp ~\vert~ \aefoldp{p} ~\vert~ \aetplp{\labelset}{\mapschema{p}{i}{\labelset}} ~\vert~ \aeinjp{\ell}{p}
\end{array}$
\end{minipage}
\caption[Syntax of the core language of $\miniVersePat$]{Syntax of the core language, $\miniVerseCore$, which is an entirely standard typed lambda calculus.  Metavariable $x$ ranges over variables, $t$ over type variables, $\ell$ over labels and $L$ over finite sets of labels. We write $\mapschema{\tau}{i}{\labelset}$ for a finite mapping of each label $i$ in $\labelset$ to some type $\tau_i$, and similarly for other sorts of terms. We write $\seqschemaX{r}$ for a finite sequence of $n > 0$  rules. %When using stylized forms, the label set is omitted when it can be inferred, e.g. the labeled product type $\prodt{\finmap{\mapitem{\ell_1}{e_1}, \mapitem{\ell_2}{e_2}}}$ leaves the label set $\{\ell_1, \ell_2\}$ implicit. 
% When we use the stylized forms, we assume that the reader can infer suppressed indices and arguments from the surrounding context.
}
\label{fig:U-expanded-terms}
\end{figure}

@@

\paragraph{Common Forms} Most of the proto-expansion forms, including all of those elided in Fig.~\ref{fig:U-candidate-terms} mirror corresponding expanded forms. The rules governing proto-expansion validation for these common forms  in the accompanying technical report correspondingly mirror the typing rules. Splicing scenes---$\tscenev$, $\escenev$ and $\pscenev$--pass opaquely through these rules, i.e. none of these rules can access the application site contexts. Notice that the initial expansion-interal typing contexts, $\Delta$ and $\Gamma$, start out empty in the rules in Fig.~\ref{fig:ee-ap-setlm}. This maintains context independence (defined formally below).

@@

\paragraph{TLM Definitions} The Reason grammar extension turns TLM definitions into module definitions of a stereotyped form. For example, the definition of \li{$regex} from Fig.~\ref{fig:regex-tlm-def} is, eliding a few minor details, implemented as shown below. The comments describe how the components correspond.

\begin{lstlisting}
    module RelitInternalDefn_regex = struct (* notation $regex *)
      type t = Regex.t (* corresponds to "at Regex.t" *)
      module Lexer_RegexLexer = struct end (* ... "lexer RegexLexer" *)
      module Parser_RegexParser = struct end (* ... "parser RegexParser" *)
      module Nonterminal_start = struct end (* ... ".start" *)
      module Package_regex_parser = struct end (* ... "in package regex_parser" *)
      module Dependencies = struct (* ... "dependencies = {" *)
        module Regex = Regex
      end
      exception Apply of string * string (* used internally, see below *)
    end
\end{lstlisting}

% The name of the TLM is stored as the module name, with the internal prefix \li{RelitInternalDefn_}. 
By encoding TLM definitions as modules and encoding the information from the TLM definition in module names, e.g. \li{Lexer_RegexLexer} and \li{Nonterminal_start}, we ensure that the corresponding signature is a \emph{singleton signature}, i.e. it uniquely identifies the TLM definition, so TLM definitions can be included in module signatures, exported from within module definitions (including those generated by functors) and packaged using the standard tooling.

@@

The Relit preprocessor rewrites TLM applications by following the specification in the previous section, differing only in that the ``dependency variable'' is a ``dependency path'', here to \li{Notation_regex___relit.Dependencies}. The only difficult is in finding the TLM definition at all, which due to our integration into the module system may have been accessed indirectly, e.g. here via the \li{open} directive and in other situations via module synonyms, functors and so on. The trick is in the singleton restriction just described---if we can determine the signature for the module path that \li{Apply} is accessed from, we have the full definition in hand. To do so, we run the OCaml typechecker on this unexpanded representation. Because the TLM application raises an exception, its type can be generalized arbitrarily. As long as unrelated type errors at a particular depth are resolved (treating the TLM applications at that depth as ``holes''), expansion at that depth can proceed. This occasionally causes types that incidentally appear in error messages to be more general than the programmer might expect (e.g. here \li{restriction_template} will have type \li{'a => 'b} during this ``pre-typing'' phase). 
%If this type ends up in an error message incidentally, e.g. due to a signature mismatch elsewhere, it may be somewhat confusing. 
We are working to improve error reporting and, more generally, improving integration with various other tools in the Reason / OCaml ecosystem.

Pattern TLMs are similar, but OCaml does not support exceptions in pattern position so we require a preprocessing step that turns any expression that the pattern binds over into an exception.

@@

Despite these semantic constraints, the system is able to express a number of non-trivial examples with few compromises because there is only one simple lexical constraint on literal bodies. The mechanism integrates cleanly into Reason/OCaml, with full support for its type system, module system and pattern matching system. TLM providers can use existing, mature parser generators and parse tree representations, and can opt-in to a stronger typing discipline by using various implementations of MetaOCaml. Overall, we believe that TLMs represent a distinctly \emph{reasonable} and \emph{expressive} new point in the design space of literal notation definition systems and macro systems.
\vspace{-5px}

@@

\clearpage
%% Bibliography
\bibliography{research}
